% !TEX root = ../master-thesis.tex

% \thispagestyle{empty}
\thispagestyle{plain}

\section*{Acknowledgements}
% \addcontentsline{toc}{section}{Acknowledgements}



I am deeply grateful to everyone who made this thesis possible through their support, collaboration, and friendship over the past years.

First and foremost, I would like to thank Philipp for being an exceptional supervisor and mentor. Your passion for physics, keen scientific insight, and support have been truly inspiring. Working with you has been an incredible journey of discovery, and I am grateful for the countless discussions that shaped both this work and my understanding of quantum many-body physics. Your ability to see the bigger picture while maintaining attention to experimental details has taught me invaluable lessons about conducting meaningful research.

I owe special thanks to the backbone of our experiment: Naman, Jin, and Marcus. Your dedication, expertise, and fearless approach to debugging the most perplexing experimental challenges have been amazing. Watching you tackle seemingly impossible problems with persistence and creativity has taught me what it truly means to be an experimental physicist. The experiment wouldn't exist without your tireless efforts, and I feel privileged to have learned from such dedicated professionals. I would also like to thank my fellow master's students Xinyi, Pragya, and Jun for their collaboration throughout this journey.

I am grateful to the broader FermiQP and Lithium teams for creating such a vibrant and collaborative research environment. The weekly discussions, lab visits, and shared experiences of building complex experiments have enriched my understanding and made the challenges more manageable.

Finally, I want to thank my family and friends for their unwavering belief in me and their constant support throughout this journey. Your encouragement has been my foundation.

This work represents not just my individual effort, but the collective contribution of an amazing community of scientists and supporters. Thank you all for making this possible.